{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce

### 2013-08-06 manuscript/bmc_article.tex
edits from kemal

Genomic structural variations (SVs), such as deletions, insertions, and inversions of DNA, are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \cite{Mills:2011p1611, Conrad:2010ja}. However, detection of SVs with current high-throughput sequencing technology remains a difficult challenge, with limited concordance between available algorithms and high false discovery rates \cite{Mills:2011p1611}. Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them to form an accurate detection measure is computationally difficult. As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure. In this paper we present a distributed software solution that is scalable and readily available on the cloud.
\par
In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data. In particular, Google's MapReduce \cite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large data sets across clusters of commodity servers. A widely adopted open-source implementation of MapReduce is provided by Hadoop, a project of the Apache Software Foundation. MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible the data they require, minimizing network traffic and I/O contention. Hadoop also provides a distributed file system (HDFS) for managing the redundant storage of large data sets across a cluster. The Hadoop framework has been shown to be effective in sequencing-related tasks including short read mapping \cite{Schatz:2009p278}, SNP calling \cite{Langmead:2009p1225}, and RNA-seq differential expression analysis \cite{Langmead:2010p1268}.

Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing problems like SV detection. MapReduce divides computation across a cluster into three phases. In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined. The framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key. Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.

% todo: keep this line?
To design MapReduce SV detection algorithms we must consider the types of variant signatures present in sequencing data sets. Popular SV detection algorithms use three main signals \cite{Alkan:2011p547}. Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \cite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation, and then clustering the discordant mappings to find SVs supported by multiple discordantly mapped read pairs. Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome \cite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints by mapping portions of individual reads to different genomic locations \cite{Wang:2011p1607,Ye:2009p2}.

@@

In the \textsc{Compute Features} job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood the alignments are correct. We then leave two functions to be implemented depending on the application:
\begin{flalign*}
 \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
 \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
\end{flalign*}

@@

SV: Structural Variation; RP: read-pair; RD: read-depth; SR: split-read; SNV: single nucleotide variations; GMM: Gaussian Mixture Model; HDFS: Hadoop File System; ROC: Receiver Operating Characteristic; FDR: False Discovery Rate; AML: acute myeloid leukemia.

@@

  \subsection*{Table 2 - Variants detected in the simulated data set}
The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses). The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section. \par \mbox{}
\par
\mbox{
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
\cline{2-9}
&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
&  BreakDancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
&  DELLY-RP           & 0.457 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
&  Pindel           & 0.462 & 0.421 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
&  MoDIL           & 0.132  & 0.66 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
   \hline
\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
\cline{2-9}
&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
&  BreakDancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
&  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
\hline
\end{tabular}
}

@@

The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class found. Exclusive predictions are in parentheses. For deletions, the same cutoffs were used as for the simulated data as in Table 3, adjusted for the difference in coverage from 30X to 37X. For insertions, the maximum sensitivity cutoff was used. \par \mbox{}
\par
\mbox{
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
  \hline
\cline{2-9}
& Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
& BreakDancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
&  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
&  DELLY-RP & 0.0931 & 0.1 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
&  Pindel & \textbf{0.179} & 0.0748 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
\hline
\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
\cline{2-9}
& Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
& BreakDancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
&  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
\hline
\end{tabular}
}

>>>>

Genomic structural variations (SVs), such as deletions, insertions, and inversions of DNA, are widely prevalent in human populations and account for the majority of the bases that differ among normal human genomes \cite{Mills:2011p1611, Conrad:2010ja}. However, detection of SVs with current high-throughput sequencing technology remains a difficult challenge, with limited concordance between available algorithms and high false discovery rates \cite{Mills:2011p1611}. Part of the problem stems from the fact that the signals indicating the presence of SVs are spread throughout large data sets, and integrating them to form an accurate detection measure is computationally difficult. As the volume of massively parallel sequencing data approaches ``big data'' scales, SV detection is becoming a time consuming component of genomics pipelines, and presents a significant challenge for research groups and clinical operations that may not be able to scale their computational infrastructure. In this paper we present a distributed software solution that is scalable and readily available on the cloud.

In other fields that have taken on the challenge of handling very large data sets, such as internet search, scalability has been addressed by computing frameworks that distribute processing to many compute nodes, each working on local copies of portions of the data. In particular, Google's MapReduce \cite{Dean:2008p277} framework was designed to manage the storage and efficient processing of very large data sets across clusters of commodity servers. A widely adopted open-source implementation of MapReduce is provided by Hadoop, a project of the Apache Software Foundation. MapReduce and Hadoop allow efficient processing of large data sets by executing tasks on nodes that are as close as possible to the data they require, minimizing network traffic and I/O contention. Hadoop also provides a distributed file system (HDFS) for managing the redundant storage of large data sets across a cluster. The Hadoop framework has been shown to be effective in sequencing-related tasks including short read mapping \cite{Schatz:2009p278}, SNP calling \cite{Langmead:2009p1225}, and RNA-seq differential expression analysis \cite{Langmead:2010p1268}.

Hadoop/MapReduce requires a specific programming model, however, which can make it difficult to design general-purpose algorithms for arbitrary sequencing data processing problems such as SV detection. MapReduce divides computation across a cluster into three phases. In the first phase, \emph{mappers} developed by the application programmer examine small blocks of data and emit a set of $\langle key, value \rangle$ pairs for each block examined. The framework then sorts the output of the mappers by key, and aggregates all values that are associated with each key. Finally, the framework executes \emph{reducers}, also created by the application developer, which process all of the values for a particular key and produce one or more outputs that summarize or aggregate those values.

% todo: keep this line?
To design MapReduce SV detection algorithms, we must consider the types of variant signatures present in sequencing data sets. Popular SV detection algorithms use three main types of signals \cite{Alkan:2011p547}. Read-pair (RP) based methods use the distance between and orientation of the mappings of paired reads to identify the signatures of SVs \cite{Campbell:2008p539,Chen:2009p3,Hormozdiari:2009p284,Sindi:2009gu,Korbel:2009dy}. Traditionally, this involves separating mappings into those that are \emph{concordant} or \emph{discordant}, where discordant mappings deviate from the expected insert size or orientation, and then clustering the discordant mappings to find SVs supported by multiple discordantly mapped read pairs. Read-depth (RD) approaches, in contrast, consider the changing depth of coverage of concordantly mapped reads along the genome \cite{Abyzov:2011bk,Alkan:2009cr,Yoon:2009kb,Chiang:2009di}. Finally, split-read (SR) methods look for breakpoints by mapping portions of individual reads to different genomic locations \cite{Wang:2011p1607,Ye:2009p2}.

@@

In the \textsc{Compute Features} job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood that the alignments are correct. We then leave two functions to be implemented depending on the application:
\begin{flalign*}
 \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
 \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
\end{flalign*}

@@

SV: Structural Variation; RP: read-pair; RD: read-depth; SR: split-read; SNV: single nucleotide variation; GMM: Gaussian Mixture Model; HDFS: Hadoop File System; ROC: Receiver Operating Characteristic; FDR: False Discovery Rate; AML: acute myeloid leukemia.

@@

  \subsection*{Table 2 - Variants Detected in the Simulated Data Set}
The number of simulated deletions and insertions in the 30X diploid chromosome 2 with Venter indels found by each tool at maximum sensitivity, as well as the number of those variants that were discovered exclusively by each tool (in parentheses). The total number of variants in each size class in the true set of deletions and insertions is shown in the first row of each section. \par \mbox{}
\par
\mbox{
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
   &                     & Prec. & Recall & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{7}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number &          &           & 224 &  84 & 82 &  31 & 26\\ 
  \hline
\cline{2-9}
&  Cloudbreak    &  0.638 & \textbf{0.678} & \textbf{153} (9)  & 61 (0) &  62 (0) & 12 (0) & 15 (0) \\ 
&  BreakDancer   &  0.356 & 0.49 & 89 (0)  & 54 (0) &  53 (0) & 8 (0) & 15 (0) \\ 
&  GASVPro        & 0.146 & 0.432 & 83 (2)  & 32 (0) &  55 (0) & 8 (0) & 15 (0) \\ 
&  DELLY-RP           & 0.457 & 0.613 & 114 (3)  & \textbf{68} (0) &  \textbf{66} (0) & 9 (1) & 17 (0) \\ 
&  DELLY-SR           & \textbf{0.679} & 0.166 & 0 (0)  & 3 (0) &  49 (0) & 6 (0) & 16 (0) \\ 
&  Pindel           & 0.462 & 0.421 & 96 (\textbf{11})  & 24 (0) &  48 (0) & 5 (0) & 15 (0)\\ 
&  MoDIL           & 0.132  & 0.66 & 123 (6)  & 66 (\textbf{3}) &  \textbf{66} (\textbf{11}) & \textbf{17} (\textbf{7}) & \textbf{23} (\textbf{8})\\ 
   \hline
\multirow{5}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number &          &           & 199 &  83 & 79 &  21 & 21\\ 
\cline{2-9}
&  Cloudbreak   &0.451 & \textbf{0.305}  & \textbf{79} (\textbf{32})  & \textbf{32} (\textbf{18}) &  \textbf{11} (8) & 1 (0) & 0 (0) \\ 
&  BreakDancer & 0.262 & 0.0968  & 23 (5)  & 14 (5) &  2 (1) & 0 (0) & 0 (0) \\ 
&  Pindel          & \textbf{0.572} & 0.196 & 52 (25)  & 5 (1) &  10 (\textbf{9}) & \textbf{3} (\textbf{2}) & \textbf{9} (\textbf{9})\\ 
&  MoDIL          & 0.186 & 0.0521 & 14 (1)  & 4 (0) &  1 (0) & 2 (\textbf{2}) & 0 (0)\\ 
\hline
\end{tabular}
}

@@

The precision and recall with respect to the gold standard set of deletions and insertions for each tool on the NA18507 data, as well as the number of variants found in each size class. Predictions exclusive to the particular algorithm are in parentheses. For deletions, the same cutoffs were used as for the simulated data as in Table 3, adjusted for the difference in coverage from 30X to 37X. For insertions, the maximum sensitivity cutoff was used. \par \mbox{}
\par
\mbox{
\begin{tabular}{r|rrr|rrrrr}
  \cline{2-9}
&  & Prec. & Recall & 40-100bp & 101-250bp & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
\hline
\multirow{6}{*}{\begin{sideways}Deletions\end{sideways}} & Total Number & & & 7,462 & 240 & 232 & 147 & 540 \\
  \hline
\cline{2-9}
& Cloudbreak & 0.0943 & \textbf{0.17} & \textbf{573} (\textbf{277})  & \textbf{176} (\textbf{30}) &  \textbf{197} (\textbf{18}) & \textbf{121} (\textbf{6}) & \textbf{399} (\textbf{24}) \\ 
& BreakDancer & 0.137 & 0.123 & 261 (29)  & 136 (3) &  178 (0) & 114 (0) & 371 (0) \\  
&  GASVPro & 0.147 & 0.0474 & 120 (21)  & 40 (2) &  85 (0) & 36 (0) & 128 (0) \\ 
&  DELLY-RP & 0.0931 & 0.1 & 143 (6)  & 128 (3) &  167 (1) & 103 (0) & 323 (1) \\ 
&  DELLY-SR & 0.153 & 0.0485 & 0 (0)  & 26 (0) &  123 (0) & 66 (0) & 203 (0) \\ 
&  Pindel & \textbf{0.179} & 0.0748 & 149 (8)  & 61 (0) &  149 (0) & 69 (1) & 217 (0) \\ 
\hline
\multirow{4}{*}{\begin{sideways}Insertions\end{sideways}} & Total Number & & & 536 & 114 & 45 & 1 & 0 \\
\cline{2-9}
& Cloudbreak & 0.0323 & \textbf{0.455} & \textbf{265} (\textbf{104})  & \textbf{49} (\textbf{24}) &  3 (1) & 0 (0)  & 0 (0)  \\ 
& BreakDancer & 0.0281 & 0.181 & 97 (10)  & 27 (5) &  2 (1) & 0 (0) & 0 (0) \\  
&  Pindel & \textbf{0.0387} & 0.239 & 144 (45)  & 14 (7) &  \textbf{7} (\textbf{6}) & \textbf{1} (\textbf{1}) &  0 (0) \\ 
\hline
\end{tabular}
}
